---
title: "A Little Book of R For Bioinformatics"
author: '@haruo_suzuki'
date: "`r Sys.time()`"
output:
  html_document:
    toc: true
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
Sys.Date()
```

```{bash}
ls -l
```

テスト用の配列データを作成する:

```{r}
# Create tests
library("seqinr")
testseq <- s2c("acgt")
length(testseq)
table(testseq)
GC(testseq)
count(testseq, wordsize = 2)
```

## [DNA Sequence Statistics (1)](http://a-little-book-of-r-for-bioinformatics.readthedocs.org/en/latest/src/chapter1.html)
**DNA配列の統計 (1)**

### [R packages for bioinformatics: Bioconductor and SeqinR](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter1.html#r-packages-for-bioinformatics-bioconductor-and-seqinr)
**バイオインフォマティクスのためのRパッケージ：BioconductorとSeqinR**

```{r}
# load the “SeqinR” R package
library("seqinr")

# ask for more information about the library() function
help("library")
```

### [Retrieving genome sequence data using SeqinR](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter1.html#retrieving-genome-sequence-data-using-seqinr)
**RパッケージSeqinRを用いて、ゲノム配列データを取得する**

```{r}
library("seqinr")
dengueseq <- read.fasta(file = "http://togows.dbcls.jp/entry/nucleotide/NC_001477.fasta")[[1]]

dengueseq[1:50]
```

### [Writing sequence data out as a FASTA file](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter1.html#writing-sequence-data-out-as-a-fasta-file)
**配列データを[FASTA](http://quma.cdb.riken.jp/help/fastaHelp_j.html)形式ファイルとして書き出す**

```{r write.fasta}
write.fasta(names="DEN-1", sequences=dengueseq, file.out="den1.fasta")
```

### [Reading sequence data into R](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter1.html#reading-sequence-data-into-r)
**配列データをRに読み込む**

```{r read.fasta}
library("seqinr")
dengue <- read.fasta(file = "den1.fasta")
dengueseq <- dengue[[1]]
```

### [Length of a DNA sequence](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter1.html#length-of-a-dna-sequence)
**DNA配列の長さ**

```{r length}
length(dengueseq)
```

### [Base composition of a DNA sequence](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter1.html#base-composition-of-a-dna-sequence)
**DNA配列の塩基組成**

```{r table}
table(dengueseq)
```

### [GC Content of DNA](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter1.html#gc-content-of-dna)
**DNAの[GC含量](https://ja.wikipedia.org/wiki/GC含量)**

```{r GC}
(2240+2770)/(3426+2240+2770+2299)
GC(dengueseq)
```

### [DNA words](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter1.html#dna-words)
**連続塩基**

```{r count}
# Count oligomers (monomer/dimer/trimer/etc)
count(seq = dengueseq, wordsize = 1)
count(seq = dengueseq, wordsize = 2)

denguetable <- count(dengueseq, wordsize = 1)
denguetable[[3]]
denguetable[["g"]]
```

## [DNA Sequence Statistics (2)](http://a-little-book-of-r-for-bioinformatics.readthedocs.org/en/latest/src/chapter2.html)
**DNA配列の統計 (2)**

### [A little more introduction to R](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter2.html#a-little-more-introduction-to-r)
**続・R言語入門**

### Reading sequence data with SeqinR

```{r}
# reading the sequence for the DEN-1 Dengue virus genome (NCBI accession: NC_001477)
library("seqinr")                          # Load the SeqinR package.
# TogoWS REST service http://togows.dbcls.jp/site/en/rest.html
dengue <- read.fasta(file = "http://togows.dbcls.jp/entry/nucleotide/NC_001477.fasta")
#dengue <- read.fasta(file = "den1.fasta") # Read in the file "den1.fasta".
dengueseq <- dengue[[1]]                   # Put the sequence in a vector called "dengueseq".

# obtain nucleotides 452-535 of DNA sequence stored in the vector `dengueseq`
dengueseq[452:535]
```

### Local variation in GC content
### [A sliding window analysis of GC content](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter2.html#a-sliding-window-analysis-of-gc-content)
GC含量の移動解析
### [A sliding window plot of GC content](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter2.html#a-sliding-window-plot-of-gc-content)
GC含量の移動プロット

```{r}
#install.packages('zoo')
library(zoo)
windowsize <- 2000
#windowsize <- 3000
#windowsize <- 300
x <- seq(from = 1, to = length(dengueseq)-windowsize, by = windowsize)
y <- rollapply(data = dengueseq, width = windowsize, by = windowsize, FUN = GC)
par(family="mono")
plot(x, y, type="b", xlab="Position (bp)", ylab="GC content")
```

### [Over-represented and under-represented DNA words](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter2.html#over-represented-and-under-represented-dna-words)
連続塩基組成 [Genomic signature](https://en.wikipedia.org/wiki/Genomic_signature) / [k-mer](https://en.wikipedia.org/wiki/K-mer)

```{r}
# Count oligomers (monomer/dimer/trimer/etc)
count(seq = dengueseq, wordsize = 2)

# Statistical over- and under- representation of oligonucleotides in a sequence
# The rho statistic can be computed on each of oligonucleotides
rho(sequence = dengueseq, wordsize = 2)
```

## [Pairwise Sequence Alignment](http://a-little-book-of-r-for-bioinformatics.readthedocs.org/en/latest/src/chapter4.html)
**[2つの配列間でのアラインメント](https://ja.wikipedia.org/wiki/シーケンスアラインメント#ペアワイズアラインメント)**


### [Retrieving a UniProt protein sequence using SeqinR](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter4.html#retrieving-a-uniprot-protein-sequence-using-seqinr)
**SeqinRでUniProtのタンパク質配列を取得**

```{r}
library("seqinr")
lepraeseq <- read.fasta(file = "http://www.uniprot.org/uniprot/Q9CD83.fasta")[[1]]
ulceransseq <- read.fasta(file = "http://www.uniprot.org/uniprot/A0PQ23.fasta")[[1]]
lepraeseq # Display the contents of "lepraeseq"
```

### [Comparing two sequences using a dotplot](http://a-little-book-of-r-for-bioinformatics.readthedocs.io/en/latest/src/chapter4.html#comparing-two-sequences-using-a-dotplot)
**ドットプロットで2つの配列を比較**

```{r}
# Create tests
library("seqinr")
testseq <- s2c("acgt")
par(family="mono"); par(mfrow=c(2,2))
dotPlot(testseq, testseq)
dotPlot(testseq, rev(testseq))
dotPlot(testseq, rep(testseq,2))
```



```{r}
# setting font in plots
par(family="mono")

dotPlot(lepraeseq, ulceransseq)
```



```{r sessionInfo}
sessionInfo()
```

### references
- https://github.com/haruosuz/DS4GD/tree/master/2019giga
- https://github.com/haruosuz/r4bioinfo/blob/master/R_Avril_Coghlan/README.md
- https://github.com/haruosuz/books/tree/master/r4ds
- https://kohske.github.io/R/rmarkdown/
